---
title: "BONCAT Activity Evaluation"
output: html_document
---

```{r "load libraries", echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
source("boncat_activity_functions.R")
opts_chunk$set(dev=c("png", "pdf"), dev.args=list(pdf = list(encoding="WinAnsi", useDingbats=FALSE)),
  fig.keep="all", fig.path = file.path("plot", "BONCAT_activity_"))
```

# Activity evaluation

What is the probability that sample from pos. > sample from neg? Use function `calculate_active_cells` in *boncat_activity_functions.R* file.

## Testing with simulated data

```{r}
# based on actual virus data parameters
load(file = file.path("data", "boncat_params.RData"))
```

Model data with a specific level of activity and see how well the activity metric does in terms of getting back out the original activity.

```{r}
# negative data set to start with
load(file = file.path("data", "boncat_data.RData"))
neg_data <- filter(boncat_all, dataset == "T7", grepl("Neg", variable))$value

# simulated result
eval_result <- expand.grid(
  approx_sim_active = c(5, 25, 50, 75, 99),
  n_particles = c(100, 1000, 10000),
  dRG_per_sub = c(0.01, 0.03, 0.05), # single substitution dRG increase, estimate from T7 (0.049)
  dRG_per_sub_range = c(0, 20, 60), # what is the +- width of the R/G signal (in percentage)
  dRG_eval_noise = c(0, 0.01), # dRG random noise (uniform distribution) for evaluating actual measurements
  avg_subs = c(1, 5, 12)  # avg number substitutions
) %>% 
  group_by_(.dots = names(.)) %>% 
  do(with(., {
    neg <- sample(neg_data, size = n_particles)
    pos <- sample(neg_data, size = n_particles)
    n_active <- floor(approx_sim_active/100*n_particles)
    # simulation data
    sim_data <- generate_activity_distribution(pos, n_active = n_active, avg_subs = avg_subs, 
                                   dRG_per_sub = dRG_per_sub, dRG_per_sub_range = dRG_per_sub_range) 
    # evaluate outcome
    active_cells <- calculate_active_cells(sim_data, neg_data, 
                                           a.noise = function(n) runif(n, -dRG_eval_noise, +dRG_eval_noise),
                                           b.noise = function(n) runif(n, -dRG_eval_noise, +dRG_eval_noise))*100
    data_frame(
      true_sim_active = n_active/n_particles*100,
      calc_active = active_cells['active'],
      lci = active_cells['lci'],
      uci = active_cells['uci']
    )
  })) %>% ungroup() 

eval_result %>% knitr::kable(d=3)
```

#### Visualize evaluation result

```{r "activity_estimate_accuracy_precision", fig.width = 10, fig.height = 8}
# plot outcome
eval_plot_df <- eval_result %>% 
  arrange(n_particles) %>% 
  mutate(
    dRG_per_sub = paste0("dRG/sub: ", dRG_per_sub),
    dRG_per_sub_range = paste0(dRG_per_sub_range, "%"),
    dRG_eval_noise = paste0("dRG noise: ", dRG_eval_noise), 
    avg_subs = sprintf("# subs: %.2d", avg_subs),
    active = sprintf("%.2d%% active",approx_sim_active)) 

base_plot <- 
  ggplot() + 
  aes(x = factor(approx_sim_active), 
      y = calc_active, shape = factor(n_particles), fill = dRG_per_sub_range, color = dRG_per_sub_range) + 
  geom_hline(data = function(df)
    df %>% select(dRG_per_sub, avg_subs, active, approx_sim_active, dRG_per_sub_range) %>% unique(),
    map = aes(yintercept = approx_sim_active, color = NULL)) +
  geom_errorbar(position=position_dodge(width = 1), map = aes(ymin = lci, ymax = uci), width = 0.5) +
  geom_point(position=position_dodge(width = 1), size = 3, color = "black") +
  labs(x = "simulated % active", y = "calculated % active", shape = "# particles measured", 
       fill = "dRG variation [+/-]", color = "dRG variation [+/-]") + 
  scale_shape_manual("# particles", values = 21:25) +
  coord_cartesian(ylim = c(0, 100)) + 
  facet_grid(avg_subs~dRG_per_sub, scales = "free_x") + 
  theme_bw() +
   guides(fill = guide_legend(override.aes = list(colour = "white", size=8, shape = 22)), 
         shape = guide_legend(override.aes = list(fill = "gray"))) 


base_plot %+% filter(eval_plot_df, grepl("0$", dRG_eval_noise)) + labs(title = "no R/G noise")
```


```{r "activity_estimate_accuracy_precision_with_noise", fig.width = 10, fig.height = 8}
base_plot %+% filter(eval_plot_df, grepl("0.01$", dRG_eval_noise)) + labs(title = "systematic R/G noise up to 0.01")
```

## Application to real data

```{r}
dRG_eval_noise <- 0.01
boncat_all %>% group_by(dataset) %>% 
  do({
    active_cells <- calculate_active_cells(filter(., grepl("Pos", variable))$value, filter(., grepl("Neg", variable))$value)*100
    active_cells_with_noise <- calculate_active_cells(
      filter(., grepl("Pos", variable))$value, filter(., grepl("Neg", variable))$value,
      a.noise = function(n) runif(n, -dRG_eval_noise, +dRG_eval_noise),
      b.noise = function(n) runif(n, -dRG_eval_noise, +dRG_eval_noise))*100
    data_frame(
      est_active = active_cells['active'],
      lci = active_cells['lci'],
      uci = active_cells['uci'],
      est_active_w_noise = active_cells_with_noise['active'],
      lci_w_noise = active_cells_with_noise['lci'],
      uci_w_noise = active_cells_with_noise['uci']
    )
  }) %>% knitr::kable(d = 2)
```

#### Further breakdown (all different replicate conditions)

```{r}
dRG_eval_noise <- 0.01
full_join(
  boncat_all %>% select(dataset, variable) %>% unique() %>% rename(var_a = variable),
  boncat_all %>% select(dataset, variable) %>% unique() %>% rename(var_b = variable),
  by = "dataset"
) %>% 
  filter(var_a != var_b) %>% 
  group_by(dataset, var_a, var_b) %>% 
  do({
    cond_a <- filter(boncat_all, dataset == .$dataset[1], variable == .$var_a[1])$value
    cond_b <- filter(boncat_all, dataset == .$dataset[1], variable == .$var_b[1])$value
    active_cells <- calculate_active_cells(cond_a, cond_b)*100
    active_cells_with_noise <- calculate_active_cells(cond_a, cond_b, 
      a.noise = function(n) runif(n, -dRG_eval_noise, +dRG_eval_noise),
      b.noise = function(n) runif(n, -dRG_eval_noise, +dRG_eval_noise))*100
    data_frame(
      est_active = active_cells['active'],
      lci = active_cells['lci'],
      uci = active_cells['uci'],
      est_active_w_noise = active_cells_with_noise['active'],
      lci_w_noise = active_cells_with_noise['lci'],
      uci_w_noise = active_cells_with_noise['uci']
    )
  })%>% knitr::kable(d = 4)
```


